---
title: 'DATA 607, Week 7: Working with HTML, XML, and JSON Files'
author: "Kevin Kirby"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview

This is the week seven for the Fall 2024 edition of DATA 607. This assignments covers the creation, importing, and manipulation of HTML, XML, and JSON files. The assignment stated:

"Pick three of your favorite books on one of your favorite subjects. At least one of the books should have more than one author. For each book, include the title, authors, and two or three other attributes that you find interesting...separately create three files which store the book’s information in HTML (using an html table), XML, and JSON formats. Create each of these files “by hand.”

In the Cursor code editor I use for my day-to-day coding, I created the three files with the following information:

* Title
* Authors
* Release Date
* Total Pages
* Objective History
  * My subjective binary on whether I believe the book is objective history

The files can be found at these links. Please note: by clicking on this like, the file will automatically download to your computer.

* [tour_books.html](https://storage.googleapis.com/data_science_masters_files/2024_fall/data_607_data_management/week_seven_files/tour_books.html)
* [tour_books.json](https://storage.googleapis.com/data_science_masters_files/2024_fall/data_607_data_management/week_seven_files/tour_books.json)
* [tour_books.xml](https://storage.googleapis.com/data_science_masters_files/2024_fall/data_607_data_management/week_seven_files/tour_books.xml)

These are these libraries required for this code to work:

```{r libraries}
library(rvest)
library(xml2)
library(jsonlite)
library(purrr)
```


## HTML file work

```{r html-manip}
gcp_html <- "https://storage.googleapis.com/data_science_masters_files/2024_fall/data_607_data_management/week_seven_files/tour_books.html"

tour_books_h <- read_html(gcp_html)
tour_books_html <- tour_books_h %>%
  html_nodes("tr") %>%
  html_nodes("td") %>%
  html_text() %>%
  matrix(ncol = 5, byrow = TRUE)
tour_books_html_df <- as.data.frame(tour_books_html, stringsAsFactors = FALSE)
colnames(tour_books_html_df) <- c("title", "authors", "release_date", "total_pages", "objective_history")

cat("HTML df:\n")
print(head(tour_books_html_df))

```


## XML file work

```{r xml-manip}
gcp_xml <- "https://storage.googleapis.com/data_science_masters_files/2024_fall/data_607_data_management/week_seven_files/tour_books.xml"

tour_books_x <- read_xml(gcp_xml)
tour_books_xml_df <- tour_books_x %>%
  xml_find_all("//book") %>%
  map_df(function(book) {
    list(
      title = xml_text(xml_find_first(book, "title")),
      authors = xml_text(xml_find_first(book, "authors")),
      release_date = xml_text(xml_find_first(book, "release_date")),
      total_pages = xml_text(xml_find_first(book, "total_pages")),
      objective_history = xml_text(xml_find_first(book, "objective_history"))
    )
  })

cat("\nXML df:\n")
print(head(tour_books_xml_df))

```


## JSON file work

Technically, I didn't need to do the as.data.frame step because json automatically comes in as a dataframe. However, my OCD kicked in and I didn't like that the page numbers in the imported dataframe structure were to the left in the cell while the othersd were to the right. Telling the dataframe to behave as a dataframe made that go away.

```{r json-manip}
gcp_json <- "https://storage.googleapis.com/data_science_masters_files/2024_fall/data_607_data_management/week_seven_files/tour_books.json"

tour_books_j <- fromJSON(gcp_json, flatten = TRUE)
tour_books_json_df <- as.data.frame(tour_books_j)

cat("\nJSON df:\n")
print(head(tour_books_json_df))


```


## Review

All three files dataframes are identical. When HTML, XML, or JSON files are imported into RStudio, R standardizes how it imports and presents structured data. Since, regardless of sematic formatting differences that support different code base usec cases, they each have key:value pairs that can be cleanly parsed into a standard table. The goal is that, no matter the data source, the data presents in a similar looking dataframe.

Where you and I may disagree is on whether JSON is structured data. I've previously heard you either say or write that you believe JSON is unustructured data (or maybe I'm completely making this up?). I view JSON as structured data that can be easily used, with minimal processing, for analytics. 

