---
title: "DATA 608 Story Two"
author: "Kevin Kirby"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

The assignment as provided is:
"The Federal Reserve’s policies and broader economic trends are often reflected in the interplay between interest rates, market prices, and bond yields. These indicators are critical for understanding economic health and investor sentiment. Your task is to analyze their relationships to determine whether patterns or trends are evident.

Research Question:

Are there visible relationships among interest rates, market prices (DJIA), and bond yields (10-year Treasury)?  
Your analysis should explore the potential interactions and correlations between these key economic indicators over the past 25 years." 

To answer these questions, I'm instructed to collect monthly data for the below from the [FRED API](https://fred.stlouisfed.org/docs/api/fred/) that's maintained by the Federal Reserve Bank of St. Louis
* Down Jones Industrial Average
* 10-Year Treasury Yield
* Unemployment Rate

Additionally, the assignment states:
"You must use base R for this assignment (e.g., plot, lines, par). External libraries such as ggplot2 are not permitted."


## API Setup

I created an account with the St. Louis Fed and received an API key. This key is stored in the `.Renviron` file as "FRED_API_KEY" and will be referenced that way below. 

```{r apiprep}
# these are baseline R libraries
library(httr)
library(jsonlite)

FRED_API_KEY <- Sys.getenv("FRED_API_KEY")
dj_id <- "DJIA"
treasury_id <- "DGS1"
unemp_id <- "UNRATE"
observ_start <- "2000-01-01"
observ_end <- "2024-12-31"
frequency <- "m"
agg_method <- "avg"
file_type <- "json"
fred_obs_url <- "https://api.stlouisfed.org/fred/series/observations"

```


### Down Jowns Industrial API Call + Cleanup 

This section performs an API call to FRED using some of the above declared parameters. 

**Please Note:** the earliest date available for this data is January 1st, 2015.

```{r djia-api}

djia_res <- GET(fred_obs_url, query = list(
  series_id = dj_id,
  api_key = FRED_API_KEY,
  observation_start = observ_start,
  observation_end = observ_end,
  frequency = frequency,
  aggregation_method = agg_method,
  file_type = file_type
))

djia_raw <- fromJSON(content(djia_res, "text"))
djia_df <- as.data.frame(djia_raw$observations)

djia_df

```

This cleans up the table to just have the month in `yyyy-mm-dd` format and the associated value. 

```{r cleanup}
djia_df <- djia_df[, c("date", "value")]
djia_df$value <- as.numeric(djia_df$value)

djia_df
```

#### 10-Year Treasury Yield API Call + Cleanup

Above, I separated API call and cleanup into separate chunks so I could explain what I was doing. Below, I've combined them into one chunk for efficiency.

**Please Note:** the earliest date available for this data is January 1st, 2000.


```{r dgs1-api}

dgs1_res <- GET(fred_obs_url, query = list(
  series_id = treasury_id,
  api_key = FRED_API_KEY,
  observation_start = observ_start,
  observation_end = observ_end,
  frequency = frequency,
  aggregation_method = agg_method,
  file_type = file_type
))

dgs1_raw <- fromJSON(content(dgs1_res, "text"))
dgs1_df <- as.data.frame(dgs1_raw$observations)

dgs1_df <- dgs1_df[, c("date", "value")]
dgs1_df$value <- as.numeric(dgs1_df$value)

dgs1_df
```

### Unemployment Rate API Call + Cleanup 

This section performs an API call to FRED using the same parameters as the above requests, with the only change needed being the `series_id`

**Please Note:** the earliest date available for this data is January 1st, 2000.

```{r unrate-api}

unrate_res <- GET(fred_obs_url, query = list(
  series_id = unemp_id,
  api_key = FRED_API_KEY,
  observation_start = observ_start,
  observation_end = observ_end,
  frequency = frequency,
  aggregation_method = agg_method,
  file_type = file_type
))

unrate_raw <- fromJSON(content(unrate_res, "text"))
unrate_df <- as.data.frame(unrate_raw$observations)

unrate_df <- unrate_df[, c("date", "value")]
unrate_df$value <- as.numeric(unrate_df$value)

unrate_df
```

## Data Merging and Cleanup

This section combines the three different dataframes into one, taking care to not drop rows when there's a null value in some but not all fields. The end result is a table that can be piped into a visualization, which is covered in the next section.

```{r date-combo}

#I checked the console and these were all characters. They need to be date format or the visualizations will have all sorts of problems. 
djia_df$date <- as.Date(djia_df$date)
dgs1_df$date <- as.Date(dgs1_df$date)
unrate_df$date <- as.Date(unrate_df$date)

#this is just changing the column names to reflect what I want them to be after the merge, otherwise I'll have three "value" fields and be lost
colnames(djia_df) <- c("month_date", "djia")
colnames(dgs1_df) <- c("month_date", "dgs1")
colnames(unrate_df) <- c("month_date", "unemp_rate")

fred_df <- merge(djia_df, dgs1_df, by = "month_date", all = TRUE)
fred_df <- merge(fred_df, unrate_df, by = "month_date", all = TRUE)

fred_df
```

## Visualizations

I spent a while trying to figure out how to complete the assignment task of producing charts for 25 years of data while also grappling with the DJIA only going back to 2015. Putting all three together on a chart where Treasury Yield and Unemployment Rate start in 2000 and the DJIA starting in 2015 created a whole lot of ugly options.

Instead, I'm producing two visualizations. Step 3 in the assignment said to create ONE visualization but the submission requirements state to submit a PDF with two. Therefore, I will create two:

1. Treasury Yield and Unemployment Rates on a monthly level from 2000 through 2024
2. All three stats starting in January 2015 and extendint through 2024


## Treasury Yield and Unemployment Rate

The graph below is a combination line and bar chart showing the generally inverse relationship between the unemployment rate and ten year treasury yields. When unemployment rises during economic downturns, like the 2008 financial crisis or the 2020 covid lock downs, the yield for ten year treasury bonds decreases. This happens because investors seek out safe havens to park their money to ride things out. The inverse happens during strong economies because the Federal Reserve is ge

The more prevalent relationship that is not part of this assignment would be assessing the relationship between 10 Year Treasury yields, unemployment rates, and Federal Reserve benchmark interest rates. 

```{r tyur}
years <- format(fred_df$month_date, "%Y")
unique_years <- unique(years)
year_labels <- unique_years[as.numeric(unique_years) %% 4 == 0]  
year_i <- match(year_labels, years)

yield_bars <- barplot(fred_df$dgs1, 
                         las = 2, 
                         col = "skyblue",
                         ylab = "Percent", 
                         main = "10 Year Treasury Yield and Unemployment Rates - 2000-2024", 
                         ylim = c(0, 16),  
                         xaxt = "n") 

lines(yield_bars, fred_df$unemp_rate, type = "l", col = "pink", lwd = 2)
text(yield_bars[year_i], par("usr")[3] - 0.5, labels = year_labels, srt = 45, adj = 1, xpd = TRUE, cex = 0.8)

legend("topleft", legend = c("Treasury Yield", "Unemployment Rate"), 
       fill = c("skyblue", NA), col = c(NA, "pink"), lty = c(NA, 1), lwd = 2)

```

### All Three Metrics

This is a combined bar and line chart showing all three metrics from January 2015 through December 2024. 

The chart shows that the stock market, as represented by the DJIA, is not an effective economic indicator. Regardless of what's happening with the Treasury Yield and Unemployment Rate, such as the stretch around the 2008 financial crisis, the stocks quickly rebound and continue marching upwards while the other metrics represent a harder slog.

```{r 3metrics}
m3_df <- fred_df[fred_df$month_date >= as.Date("2015-01-01"), ]
years_m3 <- format(m3_df$month_date, "%Y")
uy_m3 <- unique(years_m3)
yim <- match(uy_m3, years_m3)

par(mar = c(5, 5, 4, 5))

djia_bars <- barplot(m3_df$djia, beside = TRUE, 
                         ylim = c(0, max(m3_df$djia, na.rm = TRUE)), 
                         xaxt = "n",
                         las = 2, cex.names = 0.8,  
                         xlab = "Year", col = "grey")

par(new = TRUE)
plot(m3_df$month_date, m3_df$dgs1, type = "l", col = "pink", lwd = 3, 
     axes = FALSE, xlab = "", ylab = "",
     ylim = c(0, max(m3_df$dgs1, m3_df$unemp_rate, na.rm = TRUE)))

lines(m3_df$month_date, m3_df$unemp_rate, col = "blue", lwd = 3)

text(djia_bars[yim], par("usr")[3] - 0.05 * diff(par("usr")[3:4]), 
     labels = uy_m3, srt = 45, adj = 1, xpd = TRUE, cex = 0.8)

axis(4, las = 1, cex.axis = 0.8, 
     at = pretty(c(0, max(m3_df$dgs1, m3_df$unemp_rate, na.rm = TRUE))), 
     labels = formatC(pretty(c(0, max(m3_df$dgs1, m3_df$unemp_rate, na.rm = TRUE))), 
                      format = "f", digits = 2))

legend("topleft", legend = c("Dow Jones Industrial Average", "10-Year Treasury Yield", "Unemployment Rate"),
       col = c("grey", "pink", "blue"), lwd = 2, lty = c(NA, 1, 1), 
       pch = c(15, NA, NA), bty = "n")

```
