---
title: 'DATA 607 Week Nine: New York Times Newswire API'
author: "Kevin Kirby"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

This assignment asked me to make a connection with the New York Times API and take returned data and place it in a dataframe. The below does this with the style section. The API key has been hidden from view in knitting and deleted from the Rmd before submission. To run the below, please add your API key with News Wire enabled.

```{r libraries, results = FALSE, message=FALSE}
library(httr)
library(jsonlite)
library(tidyverse)

```

```{r api-hide, echo=FALSE}
nyt_key <- "add your key"
```

## Calling the. API

This is a function that calls the style section of the New York Times and returns all the content.It's designed to be basic and straight forward so that it can be later filtered.

```{r nyt-style}

fashionista <- function(section = "style") {
  vanity_fair <- "https://api.nytimes.com/svc/news/v3/content/nyt/"
  vogue <- paste0(vanity_fair, section, ".json?api-key=", nyt_key)
  
  gucci <- GET(vogue)
  
  if (status_code(gucci) == 200) {
    looks <- content(gucci, "text")
    makeup <- fromJSON(looks, flatten = TRUE)
        thoughts <- as.data.frame(makeup$results)
    
    return(thoughts)
  } else {
    message("Sad face. Nothing available.")
    return(NULL)
  }
}

style_thoughts <- fashionista("style")
style_thoughts

```

Now that I can see what the data looks like I will reduce the fields and reorder them to make more sense. 

```{r tidy-style}
fresh_looks <- style_thoughts[, c("published_date", "byline", "title", "abstract")]
fresh_looks$published_date <- as.Date(fresh_looks$published_date)
fresh_looks <- as_tibble(fresh_looks)
fresh_looks
```

## Conclusion

I treated this as a straight forward assignment with minimal need to do much beyond producing data that makes sense in a final table.

